---
title: "Running native SAINTexpress on simulated input"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running native SAINTexpress on simulated input}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
set.seed(42)
```

This vignette runs the native `SAINTexpress-spc` and `SAINTexpress-int` C++
binaries shipped by **saintexpressbin** on a small simulated AP-MS experiment.
The simulated dataset matches the one in
[`saintexpress`'s vignette](https://github.com/prolfqua/saintexpress) so you can
compare the native scores to the pure-R implementation directly.

## Check binary availability

```{r available}
library(saintexpressbin)
saintexpress_executable("spc")
saintexpress_executable("int")
saintexpress_available("spc")
```

The chunks below skip execution unless a native binary is resolvable for the
current platform.

```{r eval-flag, include = FALSE}
have_binary <- nzchar(saintexpress_executable("spc")) &&
               nzchar(saintexpress_executable("int"))
knitr::opts_chunk$set(eval = have_binary)
```

## Simulate a small AP-MS experiment

Same generator as the `saintexpress` vignette.

```{r simulate, eval = TRUE}
simulate_si <- function(seed = 42, mode = c("spc", "int")) {
  mode <- match.arg(mode)
  set.seed(seed)
  preys  <- paste0("Prey", 1:6)
  baits  <- c("BaitA", "BaitA", "BaitB", "BaitB", "Ctrl1", "Ctrl2")
  ips    <- paste0("IP", seq_along(baits))
  cort   <- c("T", "T", "T", "T", "C", "C")

  draw <- function(bait, prey) {
    if (bait %in% c("Ctrl1", "Ctrl2")) {
      if (mode == "spc") stats::rpois(1, 0.5) else stats::rexp(1, 1 / 1e4)
    } else if (bait == "BaitA" && prey %in% c("Prey1", "Prey2")) {
      if (mode == "spc") stats::rpois(1, 20) else stats::rexp(1, 1 / 1e7)
    } else if (bait == "BaitB" && prey %in% c("Prey3", "Prey4")) {
      if (mode == "spc") stats::rpois(1, 15) else stats::rexp(1, 1 / 5e6)
    } else {
      if (mode == "spc") stats::rpois(1, 0.5) else stats::rexp(1, 1 / 1e4)
    }
  }

  rows <- list()
  for (i in seq_along(baits)) {
    for (p in preys) {
      q <- draw(baits[i], p)
      if (q > 0) {
        rows[[length(rows) + 1]] <- data.frame(
          ipId = ips[i], baitId = baits[i], preyId = p, quant = q,
          stringsAsFactors = FALSE
        )
      }
    }
  }
  list(
    inter = do.call(rbind, rows),
    prey  = data.frame(preyId = preys, preyLength = 500L, preyGeneId = preys,
                       stringsAsFactors = FALSE),
    bait  = data.frame(ipId = ips, baitId = baits, CorT = cort,
                       stringsAsFactors = FALSE)
  )
}
```

## Spectral-count scoring with the native binary

```{r score-spc}
si_spc <- simulate_si(mode = "spc")
workdir <- tempfile("saintexpressbin-spc-")
dir.create(workdir)
res_spc <- saintexpress_run(si_spc, type = "spc",
                            workdir = workdir, cleanup = TRUE,
                            use_docker = FALSE)
res_spc$list[, c("Bait", "Prey", "AvgP", "BFDR", "SaintScore")]
```

Top hits per bait — the true interactors should rank first by `AvgP`:

```{r top-hits-spc}
top_per_bait <- function(df) {
  df <- df[order(df$Bait, -df$AvgP), ]
  do.call(rbind, by(df, df$Bait, head, 2))
}
top_per_bait(res_spc$list)[, c("Bait", "Prey", "AvgP", "BFDR")]
```

## Intensity scoring with the native binary

```{r score-int}
si_int <- simulate_si(mode = "int")
workdir <- tempfile("saintexpressbin-int-")
dir.create(workdir)
res_int <- saintexpress_run(si_int, type = "int",
                            workdir = workdir, cleanup = TRUE,
                            use_docker = FALSE)
top_per_bait(res_int$list)[, c("Bait", "Prey", "AvgP", "BFDR")]
```

## See also

For the pure-R implementation of the same scoring, see
[`saintexpress`](https://github.com/prolfqua/saintexpress). The two engines are
compared side by side on the TIP49 reference dataset in `prolfquasaint`.
